---
title: "DEGAS AD Experiments"
output: html_document
date: "2024-01-01"
---

```{r}
library(Seurat)
library(reticulate)
library(dplyr)
library(magrittr)
library(ggplot2)
library(GGally)
library(stringr)
library(DESeq2)
library(DEGASv2)
library(rstatix)
library(Matrix)
library(ggpubr)
library(plotly)
```


```{r}
plot_Grubman <- function(folder_path, hazard_df, sc_dataset) {
  anno_df = compare_means(hazard ~ batchCond, group.by = "cellType", data = hazard_df) %>%
   mutate(y_pos = 1.1)
  ggplot1 <- ggplot(hazard_df, aes(x=batchCond, y=hazard)) + 
    geom_boxplot(position=position_dodge()) + 
    ylim(0, 1.2) + 
    geom_point(aes(color=batchCond), position=position_jitterdodge()) +
    scale_color_manual(values = c("AD" = "darkred", "NC" = "gray"))  + 
    facet_wrap(~cellType, nrow = 2) + 
    ggsignif::geom_signif(
      data=anno_df, 
      aes(xmin=group1, xmax=group2, annotations=p.adj, y_position=y_pos), 
      manual=TRUE, textsize = 6
    ) + 
    labs(x = "Batch Condition", y = "DEGAS Hazard", color = "Batch Condition") + 
    theme(axis.title.x = element_text(face = "bold", size = 15),
          axis.title.y = element_text(face = "bold", size = 15),
          axis.text.x = element_text(face = "bold", size = 15),
          axis.text.y = element_text(face = "bold", size = 15),
          strip.text = element_text(face = "bold", size = 15),
          legend.title = element_text(face = "bold"),
          legend.text = element_text(face = "bold"))
  
  ggsave(filename = paste0(folder_path, "/box.pdf"), plot = ggplot1, width = 15, height = 6, units = "in")
  # save doublet files
  doublet_df <- hazard_df[hazard_df$cellType == "doublet", ]
  write.csv2(doublet_df, paste0(folder_path, "/doublet.csv"))
  # plot hidden features
  # phenotype <- ifelse(phenotype == 1, dis_label, NC_label)
  # plot_hidden_feat(folder_path, phenotype = phenotype, dis_label = dis_label, NC_label = NC_label) 
  # umap
  umap_data <- Embeddings(sc_dataset, "umap")
  umap_df <- data.frame(UMAP_1 = umap_data[, 1], UMAP_2 = umap_data[, 2]) %>% mutate(Hazard = hazard_df$hazard)
  ggplot(umap_df, aes(x = UMAP_1, y = UMAP_2, color = Hazard)) +
      geom_point() + 
      scale_color_gradient(low="gray", high="darkred", 
                       limits=c(0, 1)) + 
      labs(title = "UMAP Plot", x = "UMAP 1", y = "UMAP 2", color = "DEGAS Hazard")
        
  ggsave(paste0(folder_path, "/umap.pdf"), width = 10, height = 8, dpi = 300)
}
```



1. Grubman and MSBB dataset
```{r}
root_dir <- "/N/slate/zli8/DEGAS-datasets/Grubman"
raw_data <- read.csv2(paste0(root_dir, "/GSE138852_counts.csv"), header = TRUE, row.names = 1, sep = ",")
colMeta <- read.csv2(paste0(root_dir, "/GSE138852_covariates.csv"), header = TRUE, row.names = 1, sep = ",")
screateSCobj(raw_data, "Grubman", colMeta, results_save_dir = "Grubman_MSBB")
```

Load the patient data

```{r}
root_dir <- "/N/slate/zli8/DEGAS-datasets/MSBB"
bulk_dataset <- read.csv(paste0(root_dir, "/BM22_counts.csv"), header = TRUE, row.names = 1)
which(is.na(bulk_dataset)) # check no missing data
ColMeta <- read.csv(paste0(root_dir, "/BM22_meta.csv"), header = TRUE, row.names = 1)
which(colnames(bulk_dataset) != rownames(ColMeta)) # check if disease attributes alignes with patient name
phenotype_df <- data.frame(CERAD = as.integer(ColMeta$CERAD <= 2), 
                        Braak = as.integer(ColMeta$Braak >= 5),
                        CDR = as.integer(ColMeta$CDR >= 3))
```
Since there are some na values in plaquemean score, we should remove them, I also get a very balanced training dataset (88 positive versus 91 negative)
```{r}
plaque_indices <- which(!is.na(ColMeta$plaqueMean))
plaque_med <- median(ColMeta$plaqueMean, na.rm = TRUE)
reduced_bulk_dataset <- bulk_dataset[, plaque_indices]
plaque_scores <- data.frame(plaque = as.integer(ColMeta$plaqueMean[plaque_indices] > plaque_med))
```

Here is new code written in 2024

Preprocessing the Grubman and MSBB dataset
```{r}
# only run one time
# sc_dataset <- readRDS("Grubman_MSBB/SC_raw_counts/Grubman.rds")
# sc_dataset <- RunTSNE(sc_dataset, dims = 1:30)
# saveRDS(sc_dataset, "Grubman_MSBB/SC_raw_counts/Grubman_TSNE.rds")
# DEGAS_preprocessing(reduced_bulk_dataset, plaque_scores$plaque, sc_dataset, paste0("Grubman_MSBB/", "plaque"), st_lab_list = st_data@meta.data$oupSample.cellType, model_type = "categorical", diff_expr_files = c("Grubman_MSBB/SC_raw_counts/sc_diff_expr.csv"))
```

```{r}
sc_dataset <- readRDS("Grubman_MSBB/SC_raw_counts/Grubman_TSNE.rds")
```

Run the model using different feature set, see which one gives the best results
```{r}
for (feature in c("Pat_Diff", "SDE", "Pat_std")) {
  folder_path <- paste0("checkpoints/Grubman_MSBB_", "plaque", "_", feature, "_", Sys.Date())
  degas_sc_results <- run_DEGAS_SCST(paste0("Grubman_MSBB/", "plaque", "/", feature, ".RData"), "ClassClass", "Grubman_MSBB", "cross_entropy", "Wasserstein", folder_path, tot_seeds = 10)
  hazard_df <- as.data.frame(degas_sc_results[2])[, c("hazard_df.oupSample.cellType", "hazard_df.hazard", "hazard_df.oupSample.batchCond")]
  colnames(hazard_df) <- c("cellType", "hazard", "batchCond")
  hazard_df <- hazard_df %>% mutate(batchCond = ifelse(batchCond == "AD", "AD", ifelse(batchCond == "ct", "NC", batchCond)))
  write.csv2(hazard_df, paste0(folder_path, "/results.csv"))
  plot_Grubman(folder_path, hazard_df, sc_dataset)
}
```



Run model for different criterions
```{r}
for (label in c("CERAD", "Braak", "CDR")) {
  # only run one time
  # DEGAS_preprocessing(bulk_dataset, phenotype_df[, label], st_data, paste0("Grubman_MSBB/", label), st_lab_list = st_data@meta.data$oupSample.cellType, model_type = "categorical", diff_expr_files = c("Grubman_MSBB/SC_raw_counts/sc_diff_expr.csv"))
  for (feature in c("Pat_Diff")) {
    folder_path <- paste0("checkpoints/Grubman_MSBB_", label, "_", feature, "_", Sys.Date())
    degas_sc_results <- run_DEGAS_SCST(paste0("Grubman_MSBB/", label, "/", feature, ".RData"), "ClassClass", "Grubman_MSBB", "cross_entropy", "Wasserstein", folder_path, tot_seeds = 10)
    hazard_df <- as.data.frame(degas_sc_results[2])[, c("hazard_df.oupSample.cellType", "hazard_df.hazard", "hazard_df.oupSample.batchCond")]
    colnames(hazard_df) <- c("cellType", "hazard", "batchCond")
    hazard_df <- hazard_df %>% mutate(batchCond = ifelse(batchCond == "AD", "AD", ifelse(batchCond == "ct", "NC", batchCond)))
    write.csv2(hazard_df, paste0(folder_path, "/results.csv"))
    plot_Grubman(folder_path, hazard_df, sc_dataset)
  }
}
```
